---
title: "Lesson 1: Some of the Basics"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
description: >
  Gain familiarity with R by playing with some of the 
  basic operations: arithmetic, named objects, and 
  vectors.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)

gradethis_setup()

knitr::opts_chunk$set(echo = FALSE)
```


## 1. Doing Basic Arithmetic

Basic math functions are easy in R. Use the normal operators you're used to when writing out math problems, e.g. `+`, `-`, `*`, and `/`. These last two might look unusual, but they're for multiplication and division.

### 1.1 Exercises

#### 1. Add two plus two. 

In the box below, try it out by adding two and two. If you need some help, check the "Hints," and when you think you've got it, click the "Run Code" button. Finally, submit your answer for verification by clicking "Submit Answer."

```{r two-plus-two, exercise=TRUE}

```

```{r two-plus-two-solution}
2 + 2
```

```{r two-plus-two-check}
grade_this_code()
```

#### 2. Divide ten by three.

Don't stop now! Show you still have the magic in the box below.

```{r ten-divide-three, exercise=TRUE}

```

```{r ten-divide-three-solution}
10 / 3
```

```{r ten-divide-three-check}
grade_this_code()
```

## 2. Assigning names to values

In addition to basic arithmetic, R also allows you to save certain values to named objects. This can be handy for storing something in memory you can come back to later.

In R, the assignment operator looks like this: `<-`

Here's an example of it in action. Since an object's name can be almost anything, it's helpful to give it a name that's meaningful. Here, I'm assigning the value `4` to the name `legs` because that's how many legs a dog has.

```{r assignment-legs, echo=TRUE}
legs <- 4
```

This arrow is made up of the lesser-than sign and a minus sign, typed right next to each other. Keep in mind that it should be pointed left. When an object is named like this, we don't get any verification that it's happened, but we can check it by typing the name of the object to see the result:

```{r check-legs, echo=TRUE}
legs
```

### 2.1 Exercises 

#### 1. Assign a value to an object

Now you try it. In the code block below, fill in the blanks to assign the `4` to the name `legs` and the value `42` to the variable `dogs` (because that's how many dogs we apparently have).

```{r assignment-dogs, exercise=TRUE}
legs <- 4
___ ___ 42
```

```{r assignment-dogs-hint-1}
legs <- 4
dogs ___ 42
```

```{r assignment-dogs-solution}
legs <- 4
dogs <- 42
```

```{r assignment-dogs-check}
grade_this_code()
```

#### 2. Use objects in math

These named objects become most especially useful when we treat them as if they are the values themselves. Try multiplying `dogs` times `legs` in the exercise below to see how many total legs we have in our menagerie. (Remember, `legs` has the value of `4`, and `dogs` is equal to `42`.)

```{r dogs-by-legs, exercise=TRUE, exercise.setup = "assignment-dogs-solution"}

```

```{r dogs-by-legs-hint-1}
___ * ___
```

```{r dogs-by-legs-hint-2}
dogs * ___
```

```{r dogs-by-legs-solution, exercise.reveal_solution = FALSE}
dogs <- 42
legs <- 4
dogs * legs
```

```{r dogs-by-legs-check}
grade_this({
  if(identical(.result, .solution)) {
    pass("Correct, `dogs` times `legs` is the same as `42` times `4`--which is all just a roundabout way of saying `168`. You're really getting the hang of this!")
  }
  fail(random_encouragement())
})
```

## Using Vectors and Lists

Everything we've done so far is possible to do on a calculator with a memory function. R also lets us group a series of numbers together into a list, called a "vector." In its simplest form, a vector is just a series of values separated by commas.


### Exercise with Hint

*Here's an exercise where the chunk is pre-evaluated via the `exercise.eval` option (so the user can see the default output we'd like them to customize). We also add a "hint" to the correct solution via the chunk immediate below labeled `print-limit-hint`.*

Modify the following code to limit the number of rows printed to 5:

```{r print-limit, exercise=TRUE, exercise.eval=TRUE}
mtcars
```

```{r print-limit-hint}
head(mtcars)
```

### Quiz

*You can include any number of single or multiple choice questions as a quiz. Use the `question` function to define a question and the `quiz` function for grouping multiple questions together.*

Some questions to verify that you understand the purposes of various base and recommended R packages:

```{r quiz}
quiz(
  question("Which package contains functions for installing other R packages?",
    answer("base"),
    answer("tools"),
    answer("utils", correct = TRUE),
    answer("codetools")
  ),
  question("Which of the R packages listed below are used to create plots?",
    answer("lattice", correct = TRUE),
    answer("tools"),
    answer("stats"),
    answer("grid", correct = TRUE)
  )
)
```

